---
title: "Preliminary single cell analysis"
author: "Jieun Jeong"
date: "`r Sys.Date()`"
output: pdf_document
---
This example analyzes data from a single cell experiment in one of the forms
accepted by Seurat package, namely a directory with barcode.tsv, the list
of cell identifiers, genes.tsv, the list of genes with Ensembl and Symbol
rdentifiers, and matrix.mxt, counts (in 3rd column) of gene UMIs or fragments
detected for each gene (number in 1st column, row number in genes.tsv) and
each cell (number in 2nd column, row number in barcodes.tsv).

We create the initial Seurat object:
```{r, Warnings=F}
pa = "/Users/jieun/Work/Git_test/Seurat/data/filtered_gene_bc_matrices/hg19"
library(dplyr)
library(Seurat)
library(patchwork)
library(ggplot2)

pbmc.data <- Read10X(data.dir = pa)
# Initialize the Seurat object with the raw (non-normalized data).
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc
```
Quality contol metrics can be used to exclude some cells and gene from
further analysis according to the distributions of the metrics in the
data.

In the example below, we visualize QC metrics, and use these to filter cells.



```{r}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

# Visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

FeatureScatter is typically used to visualize feature-feature relationships,
but can be used for anything calculated by the object, i.e. columns in object
metadata, PC scores etc.  
```{r}
# Visualize QC metrics as a scatter plot
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```
One may use median statistics as a guide to filtering parameters.  
```{r}
MEDnF <- median(pbmc$nFeature_RNA)
MEDmt <- median(pbmc$percent.mt)
sprintf("MEDnF = %d, MEDmt = %6.4f",MEDnF,MEDmt)
```
A reasonable choice of limits on `nFeature` is at least
$0.25\times\it{MEDnF}$ and at most $3\times\it{MEDnF}$.  Similarly, one can
require that `percent.mt` is at most $2.5\times\it{MEDmt}$.  One can revise
those thresholds if the first round of the analysis provides reasons.
Improvements in technology are increasing the counts per cell and may 
require to replace parts of the workflows to fully take advantage of those
improvements, but this workflow is "traditional", predating 2022.
```{r}
pbmc <- subset(pbmc, subset = nFeature_RNA > 0.25*MEDnF & 
                 nFeature_RNA < 3*MEDnF & percent.mt < 2.5*MEDmt)
pbmc <- NormalizeData(pbmc) # log normalization and scaling
```
Improvements in technology are increasing the counts per cell and may 
require to replace parts of the workflows to fully take advantage of those
improvements, but this workflow is "traditional", predating 2022.  The next
step identifies variable features, i.e. features that help group together
similar cells and normalizes the data.
```{r}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)
# Identify most highly variable genes (illustration)
topf <- head(VariableFeatures(pbmc), 15)
# plot variable features with top labels
p <- VariableFeaturePlot(pbmc)
p <- LabelPoints(plot = p, points = topf, repel = TRUE)
p
# log normalization
allGenes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = allGenes)
```
Now we run PCA and print a highlight of the result:
```{r}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))
# Highlight: top genes in top PC dimensions
print(pbmc[["pca"]], dims = 1:5, nfeatures = 9)
```
The coefficients of PC's can be visualized too:
```{r}
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
```
We can also see a standard placement of cells in 2D diagram before clustering
```{r}
DimPlot(pbmc, reduction = "pca") + NoLegend()
```
One can also inspect how genes with high coefficient in a PC behave across cells
```{r}
DimHeatmap(pbmc, dims = 1, cells = 500, balanced = TRUE)
```
In this data sets, starting from PC_10 it is hard to see distinct patterns, so
perhaps these dimensions are not important.  While it is even more clear in Elbow Plot
below, 9 or 10 does nor need to be the best number, and Seurat designers
encourage to compare results when we use more dimensions.
```{r}
ElbowPlot(pbmc)
```
Now it is the time to prepare the main dish: the clusters!
```{r}
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)
```
We got clusters numbered from 0, and sizes can be seen simply:
```{r}
table(pbmc$seurat_clusters)
```
We can also see positions cells according to PC_1 and PC_2
```{r}
DimPlot(pbmc, reduction = "pca")
```
but usually a better view is obtained with non-linear coordinates of UMAP.  Moreover, 
after matching clusters with cell types, we can make meanigful legened with
the LabelClusters function.
```{r}
pbmc <- RunUMAP(pbmc, dims = 1:10)
DimPlot(pbmc, reduction = "umap")
```
Markers of cluster allow to identify the respective cell type base on prior
knowledge.  The function FindwMarkers() allows to find differential genes
of every individual cluster against the rest of the cells, but it also allows
to find differential genes for any pair of groups of cluster.  That may be
helpful if we want to find a classifier of all types of cells.  Of course,
it is not productive to compare every pair of cluster groups, but a figures
like two previus one hint what is worth checking.
```{r}
# find all markers of cluster 2
cluster2.markers <- FindMarkers(pbmc, ident.1 = 2)
head(cluster2.markers, n = 5)
cluster5.markers <- FindMarkers(pbmc,ident.1 = 5, ident.2 = c(1,7))
head(cluster5.markers, n = 5)
```
Some markers are common to more than one cluster, and some are "negative", so
it is interesting how they distribute among cluster.  We can find it using
three approaches: (a) violin plot of expression levels in every cluster (b) plot
of presence in UMAP diagram (c) table of summary statistic of presence in
clusters.  The last approach is best when we have many clusters, e.g. 25 or more
but the first two may allow for quick insights.  Method (c) may require a
short script, but (a) and (b) have Seurat functions.
```{r}
VlnPlot(pbmc, features = c("MS4A1", "CD79A"))
```
```{r}
VlnPlot(pbmc, features = c("MS4A1", "CD79A"), layer = "counts", log = TRUE)
```
```{r}
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A",
                               "FCGR3A", "LYZ", "PPBP","CD8A"))
```
The last plot has too many facets, we may inspect markers in smaller batches:
```{r}
FeaturePlot(pbmc, features = c("MS4A1", "GNLY", "CD3E", "CD14", "FCER1A",
                               "FCGR3A"))
```
Finally, we can have a heatmap of top markers.  To read gene names, see it in
a separate window and stretch vertically.
```{r}
# find markers for every cluster compared to all remaining cells,
# report only the positive ones
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)
pbmc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1 & -log10(p_val_adj) > 5)
pbmc.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(pbmc, features = top10$gene) + NoLegend() +
  theme(axis.text.y = element_text(size = 5))

```
We see that top 10 markers poorly differenciate between clusters 0, 2 and 4,
so we can search for alternatives using powerful subset() function of Seurat.
```{r}
pbmc_024 <- subset(pbmc, idents = c(0, 2, 4))
pbmc_024.markers <- FindAllMarkers(pbmc_024, only.pos = TRUE)
pbmc_024.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1 & -log10(p_val_adj) > 5)
pbmc_024.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 10) %>%
    ungroup() -> top10
DoHeatmap(pbmc_024, features = top10$gene) + NoLegend() +
  theme(axis.text.y = element_text(size = 7))
```  
Now cluster 4 is clearly differenciated, and for 4 vs 6 we can repeat the aproach.